---
title: "Diagnostics"
description: >
  Residual autocorrelation, aggregation standard errors, kriging adequacy, the area of applicability, and two ways a score can leak.
author: "Justin Chase"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# gstat fits the variograms behind the design effect and the kriging check, and
# ggplot2 draws the area of applicability. Both are Suggests, so the vignette is
# gated on them and R CMD build succeeds without either.
# `eval` below gates CHUNKS only: knitr does not apply it to inline `r ...`
# code, which runs regardless. Any inline reference to something a chunk
# defines has to guard itself -- `r if (runs) x else NA` -- or the vignette
# fails to build on a machine without these packages instead of skipping.
has_gstat  <- requireNamespace("gstat",   quietly = TRUE)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
runs <- has_gstat && has_ggplot

knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  out.width = "100%",
  eval      = runs
)
```

```{r missing-deps, echo=FALSE, eval=!runs, results='asis'}
cat("**gstat or ggplot2 is not installed, so the code below is shown but not",
    "run.** Install them with `install.packages(c(\"gstat\", \"ggplot2\"))`",
    "and rebuild.\n")
```

*This article needs two optional packages: **gstat**, which fits the variogram
behind the autocorrelation range, and **ggplot2** for the figures. When either
is missing the code is shown but not run, and a note at the top says so.*

## Four questions a score does not answer

A cross-validated RMSE tells you how far the predictions fell from the held-out
values. It is silent on four things that decide whether the analysis holds up:
whether the model left spatial structure in its residuals, whether the standard
errors on aggregated values are the width they claim, whether a cell mean is the
best estimate of that cell available, and where on the map the score applies at
all. A fifth section shows two ways the score itself can be wrong.

## A fixture and a model that is wrong in a known way

The response is driven by one real predictor and by a spatial field the model
never sees. `b` is noise. A linear model on `a` and `b` therefore leaves the
field in its residuals, which is what the first diagnostic should detect.

```{r fixture}
library(sf)
library(spatialkit)

set.seed(11)
n  <- 350
xy <- data.frame(x = 5e5 + runif(n, 0, 1000), y = 5e6 + runif(n, 0, 1000))
D  <- as.matrix(dist(xy))
field <- as.numeric(t(chol(exp(-D / 70) + diag(1e-8, n))) %*% rnorm(n))
xy$a  <- rnorm(n)
xy$b  <- rnorm(n)
xy$z  <- 1.5 * xy$a + field + rnorm(n, sd = 0.3)
pts   <- st_as_sf(xy, coords = c("x", "y"), crs = 32632)
```

```{r learner}
lm_fit <- function(train_sf, response_var = "z", predictor_vars = c("a", "b")) {
  d   <- st_drop_geometry(train_sf)
  fml <- as.formula(paste(response_var, "~",
                          paste(predictor_vars, collapse = " + ")))
  new_spatial_fit(
    subclass       = "lmsurf_fit",
    engine         = lm(fml, data = d),
    formula        = fml,
    response_var   = response_var,
    predictor_vars = predictor_vars,
    data_sf        = train_sf
  )
}
predict.lmsurf_fit <- function(object, newdata = NULL, ...) {
  if (is.null(newdata)) return(fitted(object$engine))
  as.numeric(predict(object$engine, newdata = st_drop_geometry(newdata)))
}
residuals.lmsurf_fit <- function(object, ...) residuals(object$engine)
fitted.lmsurf_fit    <- function(object, ...) fitted(object$engine)
for (g in c("predict", "residuals", "fitted"))
  registerS3method(g, "lmsurf_fit", get(paste0(g, ".lmsurf_fit")))

fit <- lm_fit(pts)
```

## 1. Structure left in the residuals

```{r moran}
mi <- residual_morans_i(fit)
mi
```

A large positive Moran's I says nearby residuals resemble each other, so the
model is missing something that varies smoothly across the map. Here that is the
simulated field, by construction.

Three parts of the print earn attention. The **null** says which moments the
p-value came from: `residual` moments condition on the design matrix and are
exact for OLS residuals, `randomisation` moments treat the residuals as
exchangeable and are an approximation for any model. `null = "auto"` chooses per
fit, so read what it chose. The **design rank** is the rank of the rebuilt model
matrix, which is where the residual degrees of freedom come from. The
**kurtosis** says how far the residuals sit from the Gaussian case the moments
assume, with 3 as the reference.

The weight matrix is \(n \times n\) and is described instead of returned. Ask
for it when you want to reuse the same neighbours:

```{r keep-weights}
mi_w <- residual_morans_i(fit, keep_weights = TRUE)
dim(mi_w$weights)
mi$weights_summary[c("n", "storage", "neighbours", "kept")]
```

`weights_summary` is present either way, so a `NULL` in `weights` is always
explained.

## 2. Standard errors on aggregated values

Aggregate the points into cells and the naive standard error of each cell mean
is \(s/\sqrt{n}\), which assumes the points in a cell are independent. On an
autocorrelated field they are not, and the effective sample size is smaller than
the count.

```{r cells}
hull <- clip_target_for(pts, expand = 0.02, quiet = TRUE)
tess <- build_tessellation(pts, boundary = hull, method = "square",
                           approx_n_cells = 24, quiet = TRUE)
asg  <- assign_features_to_polygons(pts, tess$cells)

c(cells = nrow(tess$cells), points = nrow(asg),
  median_per_cell = median(table(asg$poly_id)))
```

`summarize_by_cell()` takes a design effect, and three settings are worth
comparing on the same cells.

```{r deff}
sac  <- estimate_sac_range(pts, "z")
iid  <- summarize_by_cell(asg, "z", cells_sf = tess$cells, deff = 1)
kish <- summarize_by_cell(asg, "z", cells_sf = tess$cells, deff = "kish")
vgm  <- summarize_by_cell(asg, "z", cells_sf = tess$cells, deff = "variogram",
                          sac = sac)

data.frame(n         = iid$n,
           iid       = round(iid$..se_resp_z, 3),
           kish      = round(kish$..se_resp_z, 3),
           variogram = round(vgm$..se_resp_z, 3))[1:5, ]
```

```{r deff-inflation, echo=FALSE, results='asis'}
ik <- median(kish$..se_resp_z / iid$..se_resp_z, na.rm = TRUE)
iv <- median(vgm$..se_resp_z  / iid$..se_resp_z, na.rm = TRUE)
cat(sprintf(paste("The median cell's standard error is %.2f times wider under",
                  "`kish` and %.2f times wider under `variogram` than the",
                  "independent-sampling one. A confidence interval built on the",
                  "naive number is about half the width it should be on this",
                  "field.\n"), ik, iv))
```

`deff = "kish"` estimates an intra-class correlation from the within-cell and
between-cell variance and applies Kish's factor. `deff = "variogram"` takes the
correlation from the fitted variogram model, so it uses the spatial structure
rather than the cell labels. The choice is recorded:

```{r deff-attrs}
str(attr(vgm, "deff_applied"))
attr(kish, "icc")$resp
```

A cell whose design effect reaches its point count carries no information about
the variance at all, and its standard error is `NA`. That is a statement about
the cell, so the fix is bigger cells or more points, not a different setting.

## 3. Is the cell mean the best estimate of the cell?

The plain mean weights every point in a cell equally and ignores everything
outside it. Block kriging uses the fitted variogram, borrows from neighbouring
cells, and reports a variance derived from the spatial model.
`kriging_adequacy()` runs both and compares them.

```{r kriging}
ka <- kriging_adequacy(asg, "z", tess$cells, k = 4, seed = 1)
ka
```

Three readings. `kr_var` against the naive `se^2` says whether the naive
standard error was optimistic. `kr_shift` is how far the kriged estimate moved
from the plain mean, in units of the naive standard error, and a shift beyond
one or two of those says the mean and the spatial model disagree about that
cell. The blocked cross-validation at the end standardises each held-out error
by its kriging variance: a variance of 1 means the kriging variance is right,
and above 1 means it is understated.

```{r kriging-cols}
head(st_drop_geometry(ka)[, c("n", "mean", "se", "kr_pred", "kr_var", "kr_shift")], 4)
```

An empty cell still gets a kriged estimate and variance, which is the practical
reason to run this: a tessellation fine enough to be useful usually has cells
with no observation in them. This grid has none, as the report's `empty
cells` line says, because 25 cells over 350 points leaves every cell
populated; the reporting vignette's grid over the North Carolina counties has
six.

## 4. Where the score applies

A cross-validated number describes the region the training data covers. For
predictions made anywhere else, it describes nothing.
`area_of_applicability()` marks the difference, following Meyer and Pebesma
(2021).

```{r aoa}
grid <- create_grid_polygons(hull, target_cells = 200, type = "square")
gp   <- suppressWarnings(st_centroid(grid))
set.seed(2)
gp$a <- rnorm(nrow(gp))
gp$b <- rnorm(nrow(gp), mean = 3)     # shifted well outside the training range

aoa <- area_of_applicability(gp, model = fit)
aoa
```

```{r aoa-plot, fig.height=4.5, fig.alt="Cumulative distribution of the dissimilarity index. The prediction locations rise far more slowly than the training points, each measured to its nearest other training point since no folds were passed, and most of them sit beyond the dashed threshold, so the cross-validated score does not apply to most of the grid."}
plot(aoa)
```

Shifting one predictor three standard deviations moves most of the grid outside
the domain the model was trained on. The dissimilarity index `DI` is on
`aoa$aoa`, alongside the logical `AOA`, and `threshold` is the cutoff derived
from the training data.

One trap worth stating plainly. `AOA` is `NA` wherever a predictor was missing
or non-finite, `!NA` is `NA`, and R skips `NA` positions in a subscripted
assignment. Test for `TRUE`:

```{r aoa-na}
inside <- !is.na(aoa$aoa$AOA) & aoa$aoa$AOA
c(inside = sum(inside), outside_or_unknown = sum(!inside))
```

## 5. Two ways to leak

**Coordinates as predictors.** A model given `x` and `y` can reproduce the
training surface by memorising location, and random folds will not detect it.
`fit_rf_model(include_coords = TRUE)` warns for this reason. Score any such
model with blocked folds, and read `vignette("spatial-cross-validation")` for
what the blocking has to be wide enough to do.

**Selection outside the fold.** Choosing predictors on the whole data set and
then cross-validating the chosen set gives the selection a look at every test
row. `select_features_forward()` does the selection inside the resampling, and
it takes the fold scheme as an argument for the same reason the outer loop does:

```{r selection}
# select_features_forward() calls fit_fn(train_sf, vars), so the learner above
# needs an adapter that puts `vars` in the predictor slot.
sel_fn <- function(train_sf, vars) lm_fit(train_sf, "z", vars)

sel <- select_features_forward(pts, "z", c("a", "b"), fit_fn = sel_fn,
                               k = 4, method = "block_kfold", seed = 1)
sel$selected
```

Its inner folds default to the scheme you pass. Passing `random_kfold` there
produces a warning, because a leaky inner loop can select a variable for being
spatially close to the response and an honest outer loop will then report a
respectable number for a dishonestly chosen feature set.

## Next

`vignette("spatial-cross-validation")` covers fold schemes and block sizing.
`vignette("resolution")` covers choosing the cell count these diagnostics are
computed over. `?residual_morans_i` documents the two nulls and when each is
exact, `?summarize_by_cell` derives the design-effect correction and reports its
measured coverage, and `?area_of_applicability` covers the weighting and the
threshold.
