---
title: "Getting started"
description: >
  Installing, getting an sf layer of points in, what the coordinates are in, and the pipeline from points to a scored model and a map.
author: "Justin Chase"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Everything here runs on the hard dependencies and sf's own nc.shp, with two
# exceptions: the maps need ggplot2 and the model steps need ranger, both
# Suggests.  Each gated chunk leaves a note in its place when the package is
# absent, so the vignette builds either way.
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
has_ranger <- requireNamespace("ranger", quietly = TRUE)

knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## What this package is for

You have observations at points and you want values for regions. The usual move
is to aggregate onto whatever boundaries exist, and administrative boundaries
were drawn for other purposes: a census tract follows a road, a postcode follows
a delivery route. Neither follows the field you measured.

This package builds the regions from the distribution of the data, aggregates
onto them with standard errors that account for spatial dependence, and scores
models on folds that hold out whole regions. The last part matters because a
random fold on autocorrelated data returns a number that looks good and means
little.

## Installing

```{r install, eval=FALSE}
# install.packages("remotes")
remotes::install_github("elkronos/gis_modeling_toolkit")

# with the vignettes (needs pandoc)
remotes::install_github("elkronos/gis_modeling_toolkit", build_vignettes = TRUE)
```

`sf`, `dplyr`, `logger` and `digest` install with it. Everything else is
optional and is needed only for the feature that uses it:

```{r suggests-table, echo=FALSE, results='asis'}
cat(
"| install | for |\n",
"|---|---|\n",
"| `gstat` | variogram fitting: `estimate_sac_range()`, design effects, kriging checks |\n",
"| `ranger` | `fit_rf_model()` and `cv_rf()` |\n",
"| `sp`, `GWmodel` | `fit_gwr_model()` and `cv_gwr()` |\n",
"| `brms` (plus a Stan toolchain) | `fit_bayesian_spatial_model()` and `cv_bayes()` |\n",
"| `geometry` | Delaunay triangle tessellations |\n",
"| `ggplot2`, `patchwork` | every `plot_*()` function and every `plot()` method |\n",
"| `FNN`, `Matrix` | sparse k-nearest-neighbour weights for Moran's I on large layers |\n",
sep = "")
```

A missing package produces a message naming it, so nothing fails obscurely.

## Getting your data in

Your data arrives through one of two doors, and both end in the same place: an
`sf` object of POINTs in a projected CRS.

### A spatial file

```{r door-1}
library(sf)
library(spatialkit)

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
c(features = nrow(nc), geometry = as.character(unique(st_geometry_type(nc))))
```

Those are polygons, and every model here takes points, so reduce them.
`coerce_to_points()` maps a polygon to a point guaranteed to fall inside it,
using `st_point_on_surface()`, a line to its midpoint, and a MULTIPOINT to its
centroid. Attribute columns come along.

```{r door-1b}
counties <- ensure_projected(nc[, c("BIR74", "SID74", "NWBIR74")])
obs      <- coerce_to_points(counties)
obs$rate     <- 1000 * obs$SID74 / pmax(obs$BIR74, 1)   # the response
obs$nw_share <- obs$NWBIR74 / pmax(obs$BIR74, 1)        # one predictor

c(points = nrow(obs), geometry = as.character(unique(st_geometry_type(obs))))
```

Two conventions run through every function from here on. The response and
the predictors are named by their **column names**, as strings:
`response_var = "rate"`, `predictor_vars = "nw_share"`. And the response must
be numeric or logical; a factor or a character column is refused with a
message saying so.

Hex and square grids need a study area. Dissolving the source polygons is the
usual way to get one, and `st_read()` a separate outline if you have it.

```{r boundary}
boundary <- st_as_sf(st_union(counties))
```

### A table with coordinate columns

```{r door-2}
tab <- data.frame(
  lon    = st_coordinates(st_transform(obs, 4326))[, 1],
  lat    = st_coordinates(st_transform(obs, 4326))[, 2],
  births = obs$BIR74,
  sids   = obs$SID74
)
head(tab, 2)

pts_ll <- st_as_sf(tab, coords = c("lon", "lat"), crs = 4326)
pts    <- ensure_projected(pts_ll)
```

`crs =` is required and cannot be guessed. Lon/lat from a GPS, a web API or a
geocoder is almost always EPSG:4326. If the file came from a municipal or
national dataset, its metadata names the CRS, and it is frequently something
else.

## What the numbers are in

Every distance in this package is in the units of the CRS the analysis runs in.
A block size, a GWR bandwidth, a variogram range and a Gaussian process
length-scale are all distances. Hand any of them a number in degrees when the
data are in metres and the result is quietly wrong by a factor of about 100,000.

`ensure_projected()` is what prevents that. Watch what it did to the counties
above:

```{r crs}
st_crs(counties)$input
st_crs(counties)$units_gdal
```

`nc.shp` arrives in NAD27, a geographic CRS in degrees.
`ensure_projected()` chose a local Albers equal-area projection and logged why:
across this extent the UTM zone distorts distances by 0.20 percent at worst and
the Albers projection by 0.01 percent. Pass `target_crs` to override the choice.

A layer with no CRS at all is decided by a documented heuristic. If the bounding
box fits the lon/lat envelope and the coordinates look like degrees, the layer
is read as EPSG:4326 and projected, with a warning saying so. Set the CRS
yourself if your data are planar and happen to fall in that envelope.

```{r crs-missing}
raw <- st_as_sf(data.frame(lon = c(-81.5, -81.1), lat = c(36.4, 36.5)),
                coords = c("lon", "lat"))
st_crs(ensure_projected(raw))$input
```

Two layers that must line up go through `harmonize_crs()`, which returns both in
a common CRS:

```{r harmonize}
pair <- harmonize_crs(obs[1:5, ], st_transform(boundary, 4326))
st_crs(pair$a) == st_crs(pair$b)
```

You do not have to project first. Every function that measures a distance
(`make_folds()`, the `fit_*()` and `cv_*()` functions, `predict_surface()`)
calls `ensure_projected()` on what it is given, so a lon/lat layer is
projected on the way in and the same choice is logged. Calling it yourself,
as above, only means the choice is made once and can be seen.

Four more things about the data, each handled without stopping you:

- A row whose response or predictor is `NA`, `NaN` or infinite is dropped
  before a fit, and the count is logged; `prep_model_data()` is the function
  that does it and `attr(x, "dropped")` says which rows.
- Two observations at the same coordinates are fine for models and folds;
  a Voronoi tessellation keeps one seed per location
  (`build_tessellation(keep_duplicates = )` says what to do with the rest).
- A point outside the boundary you tessellate gets no cell: `tess$index`
  is `NA` for it and `assign_features_to_polygons()` leaves it out unless
  you ask for it back with `keep_unassigned = TRUE`.
- A variogram, which is what sizes blocks from the data, needs about thirty
  points; below that `estimate_sac_range()` returns `NA` and says so.

## The pipeline, end to end

Seven calls, from points to a map you can defend. The first four turn the
points into regions with aggregates; the last three fit a model on the
points, score it on folds that hold out whole regions, and say where its
predictions can be trusted. Each is covered in depth by one of the other
vignettes.

### Regions and aggregates

```{r pipeline}
# 1. How many cells can these data support?  The answer is a ranked vector,
#    best first, and the next call takes it as it is.
lv <- determine_optimal_levels(obs, max_levels = 40)
lv

# 2. Draw them.  A hex grid clipped to an irregular outline returns a few
#    more cells than asked for; params records where the count came from.
tess <- build_tessellation(obs, boundary = boundary, method = "hex",
                           approx_n_cells = lv, quiet = TRUE)

# 3. Put every observation in a cell.
asg <- assign_features_to_polygons(obs, tess$cells)

# 4. Aggregate, with standard errors corrected for within-cell correlation.
#    cells_sf attaches the geometry so the result maps; deff = "kish" is the
#    correction.
cells <- summarize_by_cell(asg, "rate", cells_sf = tess$cells, deff = "kish")

c(best_level = lv[1], cells = nrow(tess$cells))
tess$params$approx_n_cells_from
```

```{r summary-cols}
names(cells)
head(st_drop_geometry(cells)[, c("poly_id", "n", "resp_mean_rate",
                                 "..se_resp_rate")], 4)
```

Every aggregate comes with a count and a standard error. The `..sd_` and `..se_`
columns are prefixed so they cannot collide with a column of your own. A cell
with one observation has a standard error of `NA`, because one number has no
spread, and an empty cell has `n = NA`: both are kept, because a region with
nothing in it is a fact about the map.

Two ID columns can appear on a cell layer. Every tessellation carries
`cell_id`, derived from the geometry, so it is the same for the same cell
whatever order the points came in. The hex and square grids also carry
`poly_id`, which the aggregation defaults to; `assign_features_to_polygons()`
and `summarize_by_cell()` take `polygon_id_col` and `id_col` to name either,
and fall back through both when neither is named.

Eight cells, four of them slivers at the state's edge with no county inside,
is coarse. A hundred county centroids spread fairly evenly have no elbow to
find, and the call warns that `max_levels` chose this count rather than the
data. `vignette("resolution")` is where to argue with it:
`resolution_profile()` scores every cell count on four criteria at once.

```{r map, eval=has_ggplot, fig.height=4.5, fig.alt="North Carolina covered by four large hexagonal cells shaded by their mean SIDS rate, plus four unshaded slivers at the edges that hold no county. The western cell has the lowest mean, about 1.5 per thousand, and the small southern cell the highest, about 2.8."}
plot_tessellation_map(cells, boundary = boundary, fill_col = "resp_mean_rate",
                      legend_title = "mean rate")
```

```{r no-ggplot, echo=FALSE, eval=!has_ggplot, results='asis'}
cat("*(The map needs ggplot2, which is not installed here.)*\n")
```

### Cells or points?

Steps 1 to 4 answer *what is the value of each region*. The model in the next
steps is fitted on the points, because a hundred counties are a hundred
observations and eight cell means are eight. Fit on cells when the cells are
the unit you will report or act on, when the point-level data cannot leave
the building, or when the points are so dense that the cell means are what
the model should see; the reporting vignette does exactly that. Either way
the folds come from the points' locations, and the cells can serve as the
blocks that define them (`make_folds(blocks = tess$cells)`).

### A model, scored honestly

```{r folds}
# 5. Build folds that hold out whole regions.  block_size is a distance in
#    the CRS's units, here metres.  Blocks have to be at least as wide as the
#    autocorrelation range for the folds to be honest; on these data no range
#    could be identified, so 100 km is set by hand.  vignette("spatial-cross-
#    validation") is about that choice, and make_folds(auto_range = TRUE)
#    makes it from the data when gstat is installed.
random  <- make_folds(obs, k = 4, method = "random_kfold", seed = 1)
blocked <- make_folds(obs, k = 4, method = "block_kfold", block_size = 100e3,
                      seed = 1)
c(folds = blocked$k, blocks = blocked$params$blocks_used)
```

```{r cv, eval=has_ranger}
# 6. Fit and score.  cv_rf() fits a random forest inside every fold and pools
#    the held-out predictions; the same call with random folds is the number
#    a random split would have reported.
cv_random  <- cv_rf(obs, "rate", "nw_share", folds = random,  num_trees = 300, seed = 1)
cv_blocked <- cv_rf(obs, "rate", "nw_share", folds = blocked, num_trees = 300, seed = 1)
rbind(random = cv_random$overall[, c("RMSE", "MAE", "R2")],
      blocked = cv_blocked$overall[, c("RMSE", "MAE", "R2")])
```

```{r cv-folds, eval=has_ranger}
# The pooled number hides how much the folds disagree; with a hundred
# observations they disagree a lot.
rbind(random  = range(cv_random$fold_metrics$RMSE),
      blocked = range(cv_blocked$fold_metrics$RMSE))
```

```{r cv-text, echo=FALSE, eval=has_ranger, results='asis'}
ratio <- cv_blocked$overall$RMSE / cv_random$overall$RMSE
rr <- range(cv_random$fold_metrics$RMSE); rb <- range(cv_blocked$fold_metrics$RMSE)
if (ratio > 1.15) {
  cat(sprintf(paste(
    "The blocked error is %.2f times the random one: the random score was",
    "borrowing from neighbours the blocked folds deny it, which is the usual",
    "case when a model is given the coordinates or a predictor that is itself",
    "spatially smooth.\n"), ratio))
} else {
  cat(sprintf(paste(
    "The two pooled numbers differ by less than the folds do: the random",
    "folds' RMSE runs from %.2f to %.2f and the blocked folds' from %.2f to",
    "%.2f, and each pooled value sits inside the other scheme's range. On a",
    "hundred observations that is all a fold scheme can say, and it is the",
    "reason to read the per-fold range before the pooled number. The gap that",
    "blocked folds exist to expose opens when a model can borrow from its",
    "neighbours, through the coordinates or a spatially smooth predictor; the",
    "README's quick start shows one of about 1.6 on a larger simulated field.\n"),
    rr[1], rr[2], rb[1], rb[2]))
}
```

```{r cv-none, echo=FALSE, eval=!has_ranger, results='asis'}
cat("*(Steps 6 and 7 need ranger, which is not installed here. With only the",
    "hard dependencies, `cv_spatial()` scores a learner you supply through",
    "`new_spatial_fit()`; the README's quick start shows the pattern.)*\n")
```

`cv_gwr()` and `cv_bayes()` are the same call for the other two backends, and
`cv_spatial()` takes any learner you write. `compare_models_cv()` runs the
backends on the same folds and puts them in one table.

### A surface, and where to trust it

```{r surface, eval=has_ranger}
# 7. Predict on a grid over the study area and ask where the prediction is
#    extrapolating.  The area of applicability compares each grid cell's
#    predictors with the training data, using the folds above as the
#    reference, so the threshold is the one the cross-validated score was
#    earned under.
fit  <- fit_rf_model(obs, "rate", "nw_share", num_trees = 300, seed = 1)
surf <- predict_surface(fit, n_cells = 4000, covariates = obs,
                        boundary = boundary)
aoa  <- area_of_applicability(surf, model = fit, folds = blocked)
aoa
```

```{r surface-map, eval=has_ranger && has_ggplot, fig.height=4.5, fig.alt="The random forest's predicted SIDS rate on a fine grid over North Carolina, clipped to the state outline, with the hundred county points drawn over it. The surface is low across the west and centre, and highest in one south-central patch and in parts of the east, following the share of non-white births it was fitted on."}
library(ggplot2)
# predict_surface() returns the grid as points (cell centres), so the surface
# is drawn as coloured points; the county points sit on top in black.
ggplot() +
  geom_sf(data = surf, aes(colour = .pred), size = 1.1) +
  geom_sf(data = boundary, fill = NA, colour = "grey20") +
  geom_sf(data = obs, size = 0.5, colour = "black") +
  scale_colour_viridis_c(name = "predicted rate") +
  theme_void() +
  ggtitle("Predicted SIDS rate per thousand births")
```

Every grid cell is inside the area of applicability here, because the grid
took its predictor from the nearest county: nothing on it is a combination the
model has not seen. Hand `predict_surface()` a covariate raster of its own and
this is the check that says which parts of the map are guesses.

## Where to go next

```{r next-table, echo=FALSE, results='asis'}
cat(
"| vignette | covers |\n",
"|---|---|\n",
"| `vignette(\"resolution\")` | choosing the cell count, and the criteria that disagree about it |\n",
"| `vignette(\"spatial-cross-validation\")` | fold schemes, block sizing, and reading a CV result |\n",
"| `vignette(\"diagnostics\")` | residual autocorrelation, aggregation standard errors, kriging adequacy, area of applicability |\n",
"| `vignette(\"spatialkit_nc_demo\")` | the whole pipeline end to end, with maps |\n",
"| `vignette(\"reporting\")` | handing the regions to someone else, and what to report |\n",
sep = "")
```

`?spatialkit` walks the pipeline and names the function for each step. Ten
numbered scripts install with the package and run the same steps on a
simulated field with known structure, printing what to look for before each
figure:

```{r scripts, eval=FALSE}
dir <- system.file("scripts", package = "spatialkit")
list.files(dir)
Sys.setenv(SPATIALKIT_TOUR_PAUSE = "no")   # or press Enter between figures
source(file.path(dir, "03-folds.R"))
```

## Words that recur

| term | meaning here |
|---|---|
| **variogram** | semivariance of the response against the distance between pairs of points; the curve `estimate_sac_range()` fits |
| **nugget, sill, range** | the variogram's value at zero distance (noise), the value it levels off at (total variance), and the distance at which it gets there; the *effective* range is where it reaches 95 percent of the sill |
| **autocorrelation range** | that effective range: beyond it two observations are nearly independent, which is what a block or a buffer has to exceed |
| **block** | a square of the study area used to build a fold; a fold holds out whole blocks |
| **design effect (`deff`)** | how many times larger a cell mean's variance is than the independent-sample formula says, because the points in a cell are correlated; `n / deff` is the effective sample size |
| **flat region** | the cell counts a resolution criterion cannot distinguish from its optimum; a set, not an interval |
| **support ceiling, range floor** | the most cells the point count allows (`n / min_cell_n`) and the fewest the autocorrelation range allows (`area / range^2`); a criterion whose optimum sits on either is being chosen for by that bound |
| **dissimilarity index (DI)** | how far a prediction location's predictors sit from the training data, on the scale of the training data's own spread; the area of applicability is where DI stays below what cross-validation saw |
| **row ID** | `..row_id`, a column the package stamps on a layer so that folds and dropped rows can be traced back to the rows you passed in |
