---
title: "Getting started"
author: "Justin Chase"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Everything here runs on the hard dependencies and sf's own nc.shp, with one
# exception: the map at the end needs ggplot2, which is a Suggests.
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)

knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

## What this package is for

You have observations at points and you want values for regions. The usual move
is to aggregate onto whatever boundaries exist, and administrative boundaries
were drawn for other purposes: a census tract follows a road, a postcode follows
a delivery route. Neither follows the field you measured.

This package builds the regions from the distribution of the data, aggregates
onto them with standard errors that account for spatial dependence, and scores
models on folds that hold out whole regions. The last part matters because a
random fold on autocorrelated data returns a number that looks good and means
little.

## Installing

```{r install, eval=FALSE}
# install.packages("remotes")
remotes::install_github("elkronos/gis_modeling_toolkit")

# with the vignettes (needs pandoc)
remotes::install_github("elkronos/gis_modeling_toolkit", build_vignettes = TRUE)
```

`sf`, `dplyr`, `logger` and `digest` install with it. Everything else is
optional and is needed only for the feature that uses it:

```{r suggests-table, echo=FALSE, results='asis'}
cat(
"| install | for |\n",
"|---|---|\n",
"| `gstat` | variogram fitting: `estimate_sac_range()`, design effects, kriging checks |\n",
"| `ranger` | `fit_rf_model()` and `cv_rf()` |\n",
"| `sp`, `GWmodel` | `fit_gwr_model()` and `cv_gwr()` |\n",
"| `brms` (plus a Stan toolchain) | `fit_bayesian_spatial_model()` and `cv_bayes()` |\n",
"| `geometry` | Delaunay triangle tessellations |\n",
"| `ggplot2`, `patchwork` | every `plot_*()` function and every `plot()` method |\n",
"| `FNN`, `Matrix` | sparse k-nearest-neighbour weights for Moran's I on large layers |\n",
sep = "")
```

A missing package produces a message naming it, so nothing fails obscurely.

## Getting your data in

Your data arrives through one of two doors, and both end in the same place: an
`sf` object of POINTs in a projected CRS.

### A spatial file

```{r door-1}
library(sf)
library(spatialkit)

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
c(features = nrow(nc), geometry = as.character(unique(st_geometry_type(nc))))
```

Those are polygons, and every model here takes points, so reduce them.
`coerce_to_points()` maps a polygon to a point guaranteed to fall inside it,
using `st_point_on_surface()`, a line to its midpoint, and a MULTIPOINT to its
centroid. Attribute columns come along.

```{r door-1b}
counties <- ensure_projected(nc[, c("BIR74", "SID74")])
obs      <- coerce_to_points(counties)
obs$rate <- 1000 * obs$SID74 / pmax(obs$BIR74, 1)

c(points = nrow(obs), geometry = as.character(unique(st_geometry_type(obs))))
```

Hex and square grids need a study area. Dissolving the source polygons is the
usual way to get one, and `st_read()` a separate outline if you have it.

```{r boundary}
boundary <- st_as_sf(st_union(counties))
```

### A table with coordinate columns

```{r door-2}
tab <- data.frame(
  lon    = st_coordinates(st_transform(obs, 4326))[, 1],
  lat    = st_coordinates(st_transform(obs, 4326))[, 2],
  births = obs$BIR74,
  sids   = obs$SID74
)
head(tab, 2)

pts_ll <- st_as_sf(tab, coords = c("lon", "lat"), crs = 4326)
pts    <- ensure_projected(pts_ll)
```

`crs =` is required and cannot be guessed. Lon/lat from a GPS, a web API or a
geocoder is almost always EPSG:4326. If the file came from a municipal or
national dataset, its metadata names the CRS, and it is frequently something
else.

## What the numbers are in

Every distance in this package is in the units of the CRS the analysis runs in.
A block size, a GWR bandwidth, a variogram range and a Gaussian process
length-scale are all distances. Hand any of them a number in degrees when the
data are in metres and the result is quietly wrong by a factor of about 100,000.

`ensure_projected()` is what prevents that. Watch what it did to the counties
above:

```{r crs}
st_crs(counties)$input
st_crs(counties)$units_gdal
```

`nc.shp` arrives in NAD27, a geographic CRS in degrees.
`ensure_projected()` chose a local Albers equal-area projection and logged why:
across this extent the UTM zone distorts distances by 0.20 percent at worst and
the Albers projection by 0.01 percent. Pass `target_crs` to override the choice.

A layer with no CRS at all is decided by a documented heuristic. If the bounding
box fits the lon/lat envelope and the coordinates look like degrees, the layer
is read as EPSG:4326 and projected, with a warning saying so. Set the CRS
yourself if your data are planar and happen to fall in that envelope.

```{r crs-missing}
raw <- st_as_sf(data.frame(lon = c(-81.5, -81.1), lat = c(36.4, 36.5)),
                coords = c("lon", "lat"))
st_crs(ensure_projected(raw))$input
```

Two layers that must line up go through `harmonize_crs()`, which returns both in
a common CRS:

```{r harmonize}
pair <- harmonize_crs(obs[1:5, ], st_transform(boundary, 4326))
st_crs(pair$a) == st_crs(pair$b)
```

## The pipeline, end to end

Five calls, each covered in depth by one of the other vignettes.

```{r pipeline}
# 1. How many cells can these data support?
lv <- determine_optimal_levels(obs, max_levels = 40)

# 2. Draw them.
tess <- build_tessellation(obs, boundary = boundary, method = "hex",
                           approx_n_cells = 25, quiet = TRUE)

# 3. Put every observation in a cell.
asg <- assign_features_to_polygons(obs, tess$cells)

# 4. Aggregate, with standard errors corrected for within-cell correlation.
cells <- summarize_by_cell(asg, "rate", cells_sf = tess$cells, deff = "kish")

# 5. Build folds that hold out whole regions.
folds <- make_folds(obs, k = 4, method = "block_kfold", seed = 1)

c(candidate_levels = lv[1], cells = nrow(tess$cells),
  folds = folds$k, blocks = folds$params$blocks_used)
```

```{r summary-cols}
names(cells)
head(st_drop_geometry(cells)[, c("poly_id", "n", "resp_mean_rate",
                                 "..se_resp_rate")], 4)
```

Every aggregate comes with a count and a standard error. The `..sd_` and `..se_`
columns are prefixed so they cannot collide with a column of your own.

```{r map, eval=has_ggplot, fig.height=4.5}
plot_tessellation_map(cells, boundary = boundary, fill_col = "resp_mean_rate")
```

```{r no-ggplot, echo=FALSE, eval=!has_ggplot, results='asis'}
cat("*(The map needs ggplot2, which is not installed here.)*\n")
```

Cross-validating a model over those folds is `cv_spatial()` with your own
learner, or `cv_rf()`, `cv_gwr()` and `cv_bayes()` for the three backends the
package ships.

## Where to go next

```{r next-table, echo=FALSE, results='asis'}
cat(
"| vignette | covers |\n",
"|---|---|\n",
"| `vignette(\"resolution\")` | choosing the cell count, and the criteria that disagree about it |\n",
"| `vignette(\"spatial-cross-validation\")` | fold schemes, block sizing, and reading a CV result |\n",
"| `vignette(\"diagnostics\")` | residual autocorrelation, aggregation standard errors, kriging adequacy, area of applicability |\n",
"| `vignette(\"spatialkit_nc_demo\")` | the whole pipeline end to end, with maps |\n",
"| `vignette(\"reporting\")` | handing the regions to someone else, and what to report |\n",
sep = "")
```

`?spatialkit` walks the pipeline and names the function for each step.
