---
title: "What to save, and what to report"
description: >
  What leaves the session: grouping an existing layer, membership lookups, exporting the regions, and which six numbers to report.
author: "Justin Chase"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{What to save, and what to report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Everything in this vignette runs on the hard dependencies. ggplot2 draws one
# map and gstat estimates one range; both are Suggests, so those two chunks are
# gated and the rest runs anywhere.
# `eval` gates CHUNKS only: knitr does not apply it to inline `r ...` code,
# which runs regardless. Any inline reference to something a gated chunk
# defines has to guard itself -- `r if (has_gstat) x else NA` -- or the
# vignette fails to build where the package is missing instead of skipping.
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
has_gstat  <- requireNamespace("gstat",   quietly = TRUE)

knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

The other vignettes each take one decision: how many cells, which fold scheme,
which diagnostic. This one starts after those are settled. You have regions and
a validated model, and two things have to leave your session: the regions, in a
form someone else can open and join to, and the numbers that say how far the
map they are looking at can be trusted.

*Everything here runs on the hard dependencies, with two exceptions that skip
themselves with a note when absent: **ggplot2** draws the one map, and
**gstat** fits the variogram in the second half.*

## The regions as a file someone else can use

North Carolina's counties, with the 1974 SIDS rate per thousand births, and a
hexagonal grid of regions drawn from where the counties are rather than from
the county boundaries themselves. Twelve cells are asked for; a grid clipped
to an irregular outline returns more, and the count that comes back is the
one to quote.

```{r regions}
library(sf)
library(spatialkit)

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
counties <- ensure_projected(nc[, c("NAME", "BIR74", "SID74")])
counties$rate <- 1000 * counties$SID74 / counties$BIR74

centres  <- coerce_to_points(counties)
boundary <- clip_target_for(centres, expand = 0.01)
regions  <- build_tessellation(centres, method = "hex", approx_n_cells = 12,
                               boundary = boundary)$cells
nrow(regions)
```

`approx_n_cells` is approximate, as its name says: hexagons over this bounding
box came out at `r nrow(regions)` for a request of 12, and some of them will
hold no counties at all.

### Grouping a layer you already have

`assign_features_to_polygons()` is usually shown on points, but it takes
polygons too. `largest = TRUE` gives each county to the region it overlaps
most, which is what makes the result a grouping: one row per county, never a
county split across two rows.

```{r group}
grouped <- assign_features_to_polygons(counties, regions,
                                       polygon_id_col = "cell_id",
                                       largest = TRUE)

c(counties_in = nrow(counties), rows_out = nrow(grouped),
  regions_used = length(unique(stats::na.omit(grouped$cell_id))))
range(table(grouped$cell_id))
```

`r nrow(counties)` counties in, `r nrow(grouped)` rows out, spread over
`r length(unique(stats::na.omit(grouped$cell_id)))` of the `r nrow(regions)`
regions, from `r min(table(grouped$cell_id))` to
`r max(table(grouped$cell_id))` counties each. That row-count equality is the
property to check after any grouping: `sf::st_join()` would have duplicated
every county that touches two regions, and the counts per region would then be
wrong by however many counties straddle a border.

`summarize_by_cell()` reduces the grouped layer to one row per region, and
passing `cells_sf` keeps the geometry and the empty regions:

```{r aggregate}
cells <- summarize_by_cell(grouped, response_var = "rate", id_col = "cell_id",
                           cells_sf = regions, conf_level = 0.95)

names(cells)
sum(is.na(cells$n))
```

`r sum(is.na(cells$n))` of the `r nrow(cells)` regions are empty, carrying `NA`
rather than being silently dropped. A region with no data in it is a fact about
the map, and one worth keeping: it is where the model has nothing to stand on.

```{r map, eval=has_ggplot, fig.height=3.6, fig.alt="The North Carolina counties grouped into eighteen hexagonal regions, each shaded by the mean 1974 SIDS rate of the counties inside it. County outlines show underneath. Six regions along the edges hold no county and are grey."}
library(ggplot2)

ggplot() +
  geom_sf(data = cells, aes(fill = resp_mean_rate), colour = "white") +
  geom_sf(data = counties, fill = NA, colour = "grey70", linewidth = 0.2) +
  scale_fill_viridis_c(name = "SIDS rate\nper 1,000", na.value = "grey92") +
  labs(title = "Counties grouped into data-drawn regions",
       subtitle = sprintf("%d regions, %d populated; grey = no county inside",
                          nrow(cells), sum(!is.na(cells$n)))) +
  theme_void()
```

```{r no-ggplot, echo=FALSE, eval=!has_ggplot, results='asis'}
cat("*(The map needs ggplot2, which is not installed here.)*\n")
```

### Is this location inside?

The question the regions get asked most often once they leave your session is
whether some new location falls in one. That is the same function with
`keep_unassigned = TRUE`, which returns one row per location and `NA` for
anything outside every region.

```{r membership}
sites <- st_as_sf(
  data.frame(site = c("Raleigh", "Wilmington", "Atlantic", "Norfolk VA"),
             lon  = c(-78.64, -77.95, -74.50, -76.29),
             lat  = c( 35.78,  34.23,  35.00,  36.85)),
  coords = c("lon", "lat"), crs = 4326)

inside <- assign_features_to_polygons(st_transform(sites, st_crs(regions)),
                                      regions, polygon_id_col = "cell_id",
                                      keep_unassigned = TRUE)
st_drop_geometry(inside)
```

`r sum(!is.na(inside$cell_id))` of the `r nrow(inside)` sit inside a region.
The default is `keep_unassigned = FALSE`, which drops them instead; that is the
right default when you are aggregating, and the wrong one when you are
answering a membership question, because a dropped row looks like a location
you never asked about.

Two details decide whether the answer is meaningful. The locations have to be
in the regions' CRS, which is why `st_transform()` is there: the regions are in
whatever projection `ensure_projected()` chose, and lon/lat degrees compared
against metres will put everything outside. And a point on a shared border
falls in both cells, so `tie_break` decides which one it is recorded in; see
`?assign_features_to_polygons` for the rules.

### IDs that survive a reprojection

The `cell_id` a tessellation carries is already derived from the geometry, so
it does not change when the input points arrive in another order. What it is
not is portable: a file that has been reprojected, re-sorted and re-read
elsewhere has no memory of how it was built. `ensure_stable_poly_id()` derives
an ID from the geometry itself, in a common CRS, so the same region gets the
same ID from whoever computes it, in whatever projection the file is in by
then.

```{r stable-ids}
regions$src <- seq_len(nrow(regions))

from_projected <- ensure_stable_poly_id(regions[, "src"])
from_lonlat    <- ensure_stable_poly_id(st_transform(regions[, "src"], 4326))

identical(from_projected$poly_id[order(from_projected$src)],
          from_lonlat$poly_id[order(from_lonlat$src)])
```

Use that column, not the row position, as the key anything downstream joins on.

### Writing it out

The regions are an ordinary `sf` layer, so `sf::st_write()` handles them. The
format matters more than it looks:

```{r export}
dir <- tempfile(); dir.create(dir)

st_write(cells, file.path(dir, "regions.gpkg"), quiet = TRUE)
gpkg <- st_read(file.path(dir, "regions.gpkg"), quiet = TRUE)

suppressWarnings(st_write(cells, file.path(dir, "regions.shp"), quiet = TRUE))
shp <- st_read(file.path(dir, "regions.shp"), quiet = TRUE)

attr_names <- function(x) setdiff(names(x), attr(x, "sf_column"))
kept <- function(x) length(intersect(attr_names(x), attr_names(cells)))

c(columns = length(attr_names(cells)),
  GeoPackage = kept(gpkg), Shapefile = kept(shp))
```

```{r export-names, echo=FALSE}
attr_names(shp)
```

A shapefile field name is capped at ten characters, so the standard errors and
confidence bounds arrive as `X__s_rs_` and `X__c_l__`: of the
`r length(setdiff(names(cells), attr(cells, "sf_column")))` columns
`summarize_by_cell()` produced, `r length(intersect(setdiff(names(shp), attr(shp, "sf_column")), setdiff(names(cells), attr(cells, "sf_column"))))`
survive the trip. GeoPackage keeps every one of them, and renames only the
geometry column, which is a GeoPackage convention. Write a GeoPackage unless
something downstream can only read shapefiles, and if it can only read
shapefiles, rename the columns yourself first so you choose the abbreviations.

## The numbers the run already recorded

Nothing below is recomputed for the report. Each of these is written into the
object at the time the decision was made, which is the point: a report
assembled from `$params` and `$info` describes the run that happened rather
than a reconstruction of it.

The counties carry a real response, but a weak one, so the rest of this
vignette runs on a simulated field over the same county centres, with a known
correlation length of 120 km:

```{r fixture}
set.seed(42)
xy    <- st_coordinates(centres)
D     <- as.matrix(dist(xy))
field <- as.numeric(t(chol(exp(-D / 120e3) + diag(1e-8, nrow(D)))) %*%
                      rnorm(nrow(D)))

pts <- centres
pts$elev <- rnorm(nrow(pts))
pts$z    <- 0.8 * pts$elev + 2 * field + rnorm(nrow(pts), sd = 0.3)
```

```{r learner}
# A learner with no optional dependencies: a linear trend surface in the
# coordinates, plus the one covariate.
trend_fit <- function(train_sf, ...) {
  d  <- st_drop_geometry(train_sf)
  co <- st_coordinates(train_sf)
  d$x <- co[, 1]; d$y <- co[, 2]
  new_spatial_fit(subclass = "trend_fit",
                  engine = stats::lm(z ~ elev + x + y, data = d),
                  formula = z ~ elev + x + y, response_var = "z",
                  predictor_vars = "elev", data_sf = train_sf)
}
predict.trend_fit <- function(object, newdata = NULL, ...) {
  if (is.null(newdata)) return(stats::fitted(object$engine))
  d  <- st_drop_geometry(newdata)
  co <- st_coordinates(newdata)
  d$x <- co[, 1]; d$y <- co[, 2]
  as.numeric(stats::predict(object$engine, newdata = d))
}
residuals.trend_fit <- function(object, ...) stats::residuals(object$engine)
fitted.trend_fit    <- function(object, ...) stats::fitted(object$engine)
for (.g in c("predict", "residuals", "fitted"))
  registerS3method(.g, "trend_fit", get(paste0(.g, ".trend_fit")))

folds <- make_folds(pts, k = 5, method = "block_kfold", block_size = 100e3,
                    seed = 1)
cv    <- cv_spatial(pts, "z", "elev", trend_fit, folds = folds)
fit   <- trend_fit(pts)
```

## A worked report

Six items. Every one is read out of an object the run already produced.

```{r report}
m <- function(x) format(round(as.numeric(x)), big.mark = ",", scientific = FALSE)

report <- list(
  folds = sprintf("%s, k = %d, %d x %d blocks of %s m (%s)",
                  folds$method, folds$k, folds$params$grid_nx,
                  folds$params$grid_ny, m(folds$params$block_size),
                  folds$params$crs),
  score = sprintf("RMSE %.2f (per fold %.2f to %.2f), R2 %.3f",
                  cv$overall$RMSE, min(cv$fold_metrics$RMSE),
                  max(cv$fold_metrics$RMSE), cv$overall$R2),
  coverage = sprintf("%d of %d folds scored, %d rows dropped, %d orphan, %d unknown id",
                     cv$n_folds_succeeded, cv$n_folds_attempted, cv$n_dropped,
                     length(cv$orphan_rows), cv$n_unknown_ids)
)
```

```{r report-morans}
mi <- residual_morans_i(fit)
report$residuals <- sprintf("Moran's I %.3f (z = %.1f, p = %.3g, %s null)",
                            mi$observed, mi$z, mi$p_value, mi$null)
```

```{r report-aoa}
surface <- predict_surface(fit, n_cells = 1500, covariates = pts,
                           boundary = counties)
aoa <- area_of_applicability(surface, model = fit, folds = folds)
report$applicability <- sprintf("%d of %d prediction cells outside (DI > %.3f)",
                                aoa$n_outside, aoa$n_new, aoa$threshold)
```

```{r report-range, eval=has_gstat}
sac <- estimate_sac_range(pts, "z")
report$range <- sprintf("estimated range %s m against blocks of %s m",
                        m(sac), m(folds$params$block_size))
```

```{r report-print}
cat(paste0("- ", names(report), ": ", unlist(report)), sep = "\n")
```

Read together, those lines say the run failed its own checks, which is what a
report is for.

```{r block-vs-range, echo=FALSE, results='asis', eval=has_gstat}
cat(sprintf(paste0(
  "The blocks are %s m across and the correlation reaches about %s m, so ",
  "training and test data sit inside one correlation length of each other. ",
  "The score is still optimistic, and the fix is blocks at least as wide as ",
  "the range, or `auto_range = TRUE` to size them from it.\n"),
  m(folds$params$block_size), m(sac)))
```

```{r block-vs-range-nogstat, echo=FALSE, results='asis', eval=!has_gstat}
cat("The block size has to be read against the estimated autocorrelation",
    "range, which needs gstat. Blocks narrower than the range leave training",
    "and test data inside one correlation length of each other, and the score",
    "stays optimistic.\n")
```

The cross-validated R2 of `r sprintf("%.3f", cv$overall$R2)` is below zero,
which says the trend surface predicts held-out blocks worse than their own mean
does, and Moran's I of `r sprintf("%.2f", mi$observed)` on the residuals
(`r sprintf("z = %.1f", mi$z)`) says why: most of the structure in `z` is the
field, and a plane in the coordinates cannot represent it. That is the point at
which you fit something spatial rather than report this.

The applicability line reads `r aoa$n_outside` of `r aoa$n_new` here because
`predict_surface()` joined the covariate from the nearest observation, so no
grid cell holds a combination the model never saw. On a real covariate raster
it will not be zero; `vignette("diagnostics")` works through what to do when it
is not.

## What each line is for

```{r report-table, echo=FALSE, results='asis'}
cat(
"| Report | From | Because |\n",
"|---|---|---|\n",
"| The fold scheme, the block size and the estimated range, in named CRS units | `folds$params` | A cross-validation score means nothing without the geometry that produced it: blocks below the range leak, and the score comes back flattering (Roberts et al. 2017) |\n",
"| The metric **with its per-fold spread**, not the pooled value alone | `cv$fold_metrics`, `plot_cv_metrics()` | 0.9 in four folds and 14 in one is a model that fails in one region, and the pooled number cannot tell that from uniform mediocrity |\n",
"| Residual Moran's I with its null | `residual_morans_i()` | It is the check on whether the spatial structure the model was built for actually got modelled |\n",
"| The share of the prediction area outside the area of applicability | `area_of_applicability()` | A fitted model returns a number everywhere, including ground nothing like its training data, and the CV score does not apply there (Meyer and Pebesma 2021) |\n",
"| Interval coverage, for a model with intervals | `cv_bayes()`, `plot_calibration()` | Good point predictions routinely come with badly calibrated intervals, so RMSE alone can prefer the model whose uncertainty is wrong (Heaton et al. 2019) |\n",
"| Rows dropped, folds that failed, orphan and unknown IDs | `cv$n_dropped`, `cv$orphan_rows`, `cv$fold_status` | A score computed over 80 percent of the data is not a score over the data, and nothing else in the output says so |\n",
sep = "")
```

Two of the six are about what the number does **not** cover, which is the part
that gets left out. The area of applicability bounds the map in space and the
dropped-row counts bound it in sample: without them a reader has a figure and
no idea what it was computed over.

For the aggregates themselves, report the standard error that comes with each
cell mean rather than the mean alone. `summarize_by_cell()` computes it with
the observation count beside it, and `deff` corrects it for within-cell
autocorrelation. A cell mean from two observations and one from thirteen are
different claims, and printed side by side they look identical.

```{r se-example}
populated <- cells[!is.na(cells$n), ]
summary(populated$n)
range(populated$..se_resp_rate, na.rm = TRUE)
```

## What to save

Enough to rerun it, which is less than it sounds:

- the seed (`folds$params$seed`) and the fold assignment (`folds$assignment`),
- the regions, with `ensure_stable_poly_id()` applied, as a GeoPackage,
- `folds$params` and the fit's `$info`, which hold every decision made for you,
- the session's package versions.

The regions are the one output that cannot be regenerated from the others
without care. `get_voronoi_seeds()` and the k-means sweep in
`determine_optimal_levels()` are seeded, so they repeat, but only at the same
seed and the same package version. Save the layer.

## Next

`vignette("getting-started")` is the shortest path through the pipeline that
produced all of this. `vignette("diagnostics")` covers the residual and
applicability checks the table asks you to report, and
`vignette("spatial-cross-validation")` covers fold geometry and how to read a
CV result down to the last row.

## References

Heaton, M. J., Datta, A., Finley, A. O., et al. (2019). A case study
competition among methods for analyzing large spatial data. *Journal of
Agricultural, Biological and Environmental Statistics* 24, 398-425.
<https://doi.org/10.1007/s13253-018-00348-w>

Meyer, H. and Pebesma, E. (2021). Predicting into unknown space? Estimating
the area of applicability of spatial prediction models. *Methods in Ecology
and Evolution* 12(9), 1620-1633.
<https://doi.org/10.1111/2041-210X.13650>

Roberts, D. R., Bahn, V., Ciuti, S., et al. (2017). Cross-validation
strategies for data with temporal, spatial, hierarchical, or phylogenetic
structure. *Ecography* 40(8), 913-929.
<https://doi.org/10.1111/ecog.02881>

```{r session, echo=FALSE}
sessionInfo()
```
