---
title: "Choosing a resolution"
description: >
  How many cells the data support: the ladder, four criteria that disagree, and how to read the flat region.
author: "Justin Chase"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Choosing a resolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# gstat supplies the variogram behind two of the four criteria, and ggplot2
# draws the profile. Both are Suggests, so the vignette is gated on them here
# and R CMD build succeeds on a machine that has neither.
# `eval` below gates CHUNKS only: knitr does not apply it to inline `r ...`
# code, which runs regardless. Any inline reference to something a chunk
# defines has to guard itself -- `r if (runs) x else NA` -- or the vignette
# fails to build on a machine without these packages instead of skipping.
has_gstat  <- requireNamespace("gstat",   quietly = TRUE)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
runs <- has_gstat && has_ggplot

knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  out.width = "100%",
  eval      = runs
)
```

```{r missing-deps, echo=FALSE, eval=!runs, results='asis'}
cat("**gstat or ggplot2 is not installed, so the code below is shown but not",
    "run.** Install them with `install.packages(c(\"gstat\", \"ggplot2\"))`",
    "and rebuild.\n")
```

*This article needs two optional packages: **gstat**, which fits the variogram
behind the autocorrelation range, and **ggplot2** for the figures. When either
is missing the code is shown but not run, and a note at the top says so.*

## The question

Before point observations can be aggregated into regions, something has to
decide how many regions. Twenty cells over a county smooth away most of the
variation you came to measure. Two thousand give you cells holding one
observation each, where the standard errors are undefined and every cell mean is
one number pretending to be an average.

Administrative boundaries settle this by fiat: you get the census tracts that
exist. Drawing the regions yourself removes the fiat and leaves the decision
with you, and the number propagates into the variance of the cell means, the
design effect, and the size of the blocks a spatial cross-validation holds out.

Three functions answer the question, in increasing order of how much they tell
you. `determine_optimal_levels()` reads an elbow off the coordinates alone and
runs on the hard dependencies; it is the quick answer and the one
`build_tessellation()` accepts directly. `resolution_profile()` scores every
cell count on a ladder against four criteria at once and needs gstat for two
of them. `select_resolution()` reads one criterion off that profile, and
`summary()` on the profile reads all four side by side. If the cells will feed
a model, the profile is the one to use. If you need a count now and the data
are all you have, the elbow is defensible when the points cluster, and this
article says where it falls short. On points spread evenly there is no elbow
to read, and both functions say so rather than hand back a count the ladder's
ends chose.

The argument that says "how many" is spelled differently by the function it
belongs to: `max_levels` bounds the elbow's ladder, `n_levels` sets the
profile's, `approx_n_cells` and `target_cells` are what the grid builders
take, and `n` is what `get_voronoi_seeds()` takes. Each of the last three
accepts the object the first two return.

## A fixture with known structure

A simulated exponential field, so the structure is known before anything is
fitted. The covariance parameter `a = 100` gives an effective range of 300 units
across a 1000-unit square, and `sd = 1` adds a nugget large enough to matter
later.

```{r fixture}
library(sf)
library(spatialkit)

set.seed(42)
n  <- 500
xy <- data.frame(x = 5e5 + runif(n, 0, 1000), y = 5e6 + runif(n, 0, 1000))
D  <- as.matrix(dist(xy))
xy$z <- as.numeric(t(chol(exp(-D / 100) + diag(1e-8, n))) %*% rnorm(n)) +
        rnorm(n, sd = 1)
pts  <- st_as_sf(xy, coords = c("x", "y"), crs = 32632)
```

## The short answer

`determine_optimal_levels()` returns a few candidate cell counts, best first.

```{r doi}
lv <- determine_optimal_levels(pts, max_levels = 40)
lv
```

This fixture's points are uniform, so their within-cluster sum of squares falls
like `c / k` all the way down and bends nowhere of its own. The call warns that
it found no elbow and that `max_levels` chose `lv`: a larger `max_levels` would
give a larger answer. Read an elbow only off points that cluster; here the
profile below is the better guide.

`lv[1]` is what `get_voronoi_seeds()` takes as `n`, and what
`build_tessellation()` takes as `approx_n_cells` for its `"hex"` and `"square"`
methods. Voronoi has no cell count to set: it grows one cell per point you hand
it, which is why the count is set when you place the seeds rather than when you
tessellate them. The last section shows the two calls in order.

Two things to know about this function before you rely on it. Its
`criterion = "morans_i"` and `criterion = "combined"` settings need both
`response_var` and `predictor_vars`; give it only a response and it warns
and falls back to `"geometric"`. And its ladder runs from 1 to
`max_levels` with no lower bound from the spatial correlation of the data, so it
can prefer cells wider than the field's own correlation range. The profile below
does impose that bound, which is why the two can disagree.

## The profile

`resolution_profile()` fits every level on a ladder and reports what each one
costs, so you can see the shape of the trade-off before picking a point on it.
Levels are spaced logarithmically, because cell diameter falls as
\(L^{-1/2}\).

```{r profile}
prof <- resolution_profile(pts, response_var = "z", n_levels = 16)
prof
```

One row per level. `cell_n_median` and `cell_diam_median` are usually the first
two columns worth reading: how many points a typical cell holds, and how big it
is in CRS units. `cell_diam_median` is twice the median root-mean-square
distance of a cell's points from its centre, about 0.8 of the side of a square
cell of the same area, so compare it against something you already know about
your own data, such as the spacing of a sampling grid or the size of a field,
with that factor in mind.

### The four criteria

```{r criteria-table, echo=FALSE, results='asis'}
cat(
"| criterion | measures | needs | direction |\n",
"|---|---|---|---|\n",
"| `elbow` | how far the within-cluster sum of squares curve bends below its own chord | coordinates only | larger |\n",
"| `cp` | Mallows' \\(C_p\\) of the piecewise-constant approximation of the response by cell means | a response, and a variogram for the nugget | smaller |\n",
"| `moran_z` | spatial structure surviving in the residuals of the cell means | a response | nearer zero |\n",
"| `reliability` | the share of the spread in the cell means that is signal, not sampling noise | a variogram | larger |\n",
sep = "")
```

`elbow` sees the coordinates and knows nothing about what you measured. `cp` and
`reliability` split the practical question in half: how well the cells represent
the field, and whether the cell values can be told apart from noise. `moran_z`
is the diagnostic, and a large deviate says the cells are still leaving spatial
pattern on the table.

```{r plot-profile, fig.height=8, fig.alt="Three stacked panels, one per criterion the profile could score, against the number of cells on a shared axis; there is no elbow panel because these uniform points have no elbow. Mallows' Cp falls from 10 to 49 cells and turns up at 55, reliability declines steadily from 10 cells on, and the absolute Moran's z is lowest at 20 cells and climbs past 30. A red dot and dotted line mark each criterion's choice, and shaded bands the levels within tolerance of it. The bands are one to three levels wide, none has a gap, and no level sits inside all three."}
plot(prof)
```

Two behaviours are worth knowing before reading the numbers.

```{r cp-context, echo=FALSE}
vg  <- attr(prof, "variogram")
nug <- round(vg$nugget, 2); psl <- round(vg$psill, 2)
```

**`cp` needs a nugget to have an interior optimum.** On a smooth field the
piecewise-constant approximation keeps improving as cells shrink, the penalty is
too small to stop it, and the minimum lands wherever `min_cell_n` stops the
ladder. The support ceiling is then doing the choosing, and
`select_resolution()` flags it. This fixture has a fitted nugget of
`r if (runs) nug else NA` against a partial sill of `r if (runs) psl else NA`,
which is why `cp` turns around inside the ladder here.

**`reliability` rises as cells get larger**, because bigger cells hold more
points and average away more noise. Its optimum can therefore sit on the coarse
end of whatever the ladder allows, which makes the answer a bound on the
analysis. The next section is about spotting that.

## Choosing, and the flat region

`select_resolution()` picks the level a criterion prefers and reports the band
around it.

```{r select}
sel <- select_resolution(prof, criterion = "reliability")
sel
```

`sel$best` is the optimum and `sel$flat` is every level within `tol` of it.
Quote the band when you write the analysis up. `tol` is relative to the
optimum's value for `cp` and `reliability`, and relative to the criterion's
range for `elbow` and `moran_z`, whose optima can be zero.

Two flags say when a bound chose instead of the criterion:

```{r flags}
c(at_floor = sel$at_floor, at_ceiling = sel$at_ceiling)
```

```{r flag-text, echo=FALSE, results='asis'}
if (isTRUE(sel$at_floor)) {
  cat("`at_floor` is `TRUE` here. Reliability wanted to keep going coarser and",
      "the correlation-range floor stopped it, so read the answer as *at most",
      "this fine* and take the number as a bound on the analysis, not as an",
      "optimum. The floor is a fact about the data, so the fix is a different",
      "question, not a different `tol`.\n")
} else {
  cat("Both are `FALSE` here, so the optimum sits inside the ladder and the",
      "criterion is the thing that chose it.\n")
}
```

Criteria can prefer different levels while agreeing on a region. `summary()`
on the profile reads all four at once, and closes with the levels that every
band contains:

```{r compare-criteria}
summary(prof)
```

```{r band-text, echo=FALSE, results='asis'}
# A band is a set: the curves are not monotone, so a band can skip rungs. The
# sentence below reads the rendered summary rather than quoting numbers that
# a different platform's k-means or variogram fit could move.
s     <- summary(prof)
lad   <- attr(s, "ladder")
bands <- attr(s, "bands")[s$criterion]
pos   <- lapply(bands, function(b) sort(match(b, lad)))
holed <- names(pos)[vapply(pos, function(p) any(diff(p) > 1L), logical(1))]
solid <- names(pos)[vapply(pos, function(p) length(p) > 1L && all(diff(p) == 1L),
                           logical(1))]
oxford <- function(x) if (length(x) < 2L) as.character(x) else
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
example <- if (length(holed)) {
  p <- pos[[holed[1L]]]
  skipped <- lad[setdiff(seq(min(p), max(p)), p)]
  sprintf("the `%s` band above accepts %s and rejects the %s between them,",
          holed[1L], oxford(lad[p]), oxford(skipped))
} else {
  "a band that skips a rung prints as a comma-separated list,"
}
run <- if (length(solid)) {
  p <- pos[[solid[1L]]]
  sprintf("while a solid run of %d rungs prints as a range, as `%s` does with `%d to %d`.",
          length(p), solid[1L], lad[min(p)], lad[max(p)])
} else "while a solid run prints as a range."
cat("A band is a set of levels, not an interval, because the criterion curves",
    "are not monotone:", example, run,
    "Where the bands overlap you have a defensible set of levels, and the last",
    "line names it; `attr(summary(prof), \"common\")` returns the same levels",
    "for use in code, and `attr(summary(prof), \"bands\")` each criterion's",
    "region in full. Where they do not overlap, the criteria are answering",
    "different questions and you have to say which one your analysis needs.",
    "An empty intersection is a result: it says this field has no single",
    "resolution that satisfies every way of asking.\n")
```

The table is not a decision procedure, and nothing in the package will pick
for you. Cross-validating a model at each suggested level is affordable, but
it does not settle the question either: on a simulated field the level that
won on cross-validated \(R^2\) moved with the fold seed, and the coarsest
grid won most often because its score had the widest spread, not because it
was better.

## What the ladder can support

The ladder runs between two bounds the data impose, both recorded on the profile.

```{r bounds}
str(attr(prof, "bounds"))
```

The ceiling is `floor(n / min_cell_n)`, or the number of distinct locations
when that is smaller (one short of it when no location repeats); `ceiling_from`
says which. Past it the average cell holds too few points to estimate anything
from, or k-means has more centres to place than distinct points. The floor is `ceiling(area / range^2)`, from
the fitted autocorrelation range: cells wider than the range average over more
than one patch of the field, mixing values the field itself keeps apart.

When the floor exceeds the ceiling, the data cannot support a tessellation that
respects their own correlation structure. `supported` is `FALSE`, a warning says
so, and the ladder still runs from 2 to the ceiling so the cost of each level
stays visible. The usual causes are too few points for the extent, or a
correlation range shorter than the spacing between observations.

## Choosing on one half, estimating on the other

Picking the level from the same response you then aggregate is a form of
double-dipping: the cell count was chosen to suit one realisation of the field.
`select_on = "split"` profiles part of the layer and hands back the row positions
of both parts.

```{r split}
prof_split <- resolution_profile(pts, response_var = "z", n_levels = 16,
                                 select_on = "split")
sp <- attr(prof_split, "split")
sp
```

The split is spatially blocked, not random, and records the method and seed that
produced it. A random half would put neighbours of every selection point in the
estimation set, and on an autocorrelated field that leaks the structure you are
choosing against.

Choose from `prof_split`, then aggregate the rows in `sp$estimation`. The cells
on the ladder are still drawn on every point, so the count it picks is a count
for the whole layer, and the ladder is as long as the full profile's. Only the
steps that read the response (the variogram, `cp`, `moran_z`) use the
selection half. The cost is power: those criteria see half the data, so the
profile is noisier and the band wider. The estimation rows also fill only
their own half of the layer: a cell inside the selection half gets none of
them, and a cell across the border between the halves is estimated from the
part of it on the estimation side. Read standard errors only off cells whose
points are all estimation rows.

```{r split-ladder}
unlist(attr(prof_split, "bounds")[c("floor", "ceiling", "n", "supported")])
```

The floor can differ from the full profile's, because it comes from the range
estimated on the selection half.

## Next

Hand the chosen number to whichever call decides the cell count. For a Voronoi
tessellation that is `get_voronoi_seeds()`, and `build_tessellation()` then works
on the seeds:

```{r next-step}
bnd   <- clip_target_for(pts, expand = 0.02, quiet = TRUE)
seeds <- get_voronoi_seeds(boundary = bnd, method = "kmeans", n = sel$best,
                           sample_points = pts, set_seed = 1)
cells <- build_tessellation(seeds, boundary = bnd, method = "voronoi",
                            quiet = TRUE)
nrow(cells$cells)
```

For a lattice the count goes to `build_tessellation()` itself, and the points go
in directly:

```{r next-step-grid}
hex <- build_tessellation(pts, boundary = bnd, method = "hex",
                          approx_n_cells = sel$best, quiet = TRUE)
nrow(hex$cells)
```

The hex count overshoots the request: the target is adjusted for packing density
and then clipped to an irregular boundary, so `approx_n_cells` is approximate in
both directions. The seeded Voronoi hits the number exactly, because the seeds
are the cells.

Passing `approx_n_cells` with `method = "voronoi"` warns and is ignored. Without
the seeding step you get one cell per observation, which is a nearest-neighbour
interpolation rather than a resolution.

`vignette("spatialkit_nc_demo")` runs the whole pipeline on real boundaries.
`?resolution_profile` documents how each criterion behaved on simulated fields,
with the references; `?determine_optimal_levels` covers the k-means++ restart
budget and the nine-cell floor below which the model-aware criteria carry no
information.
