---
title: "Spatial cross-validation"
description: >
  The five fold schemes, sizing blocks from the autocorrelation range, and reading a cross-validation result down to the last row.
author: "Justin Chase"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Spatial cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# gstat estimates the autocorrelation range that sizes the blocks, and ggplot2
# draws the folds and the sweep. Both are Suggests, so the vignette is gated on
# them and R CMD build succeeds without either.
# `eval` below gates CHUNKS only: knitr does not apply it to inline `r ...`
# code, which runs regardless. Any inline reference to something a chunk
# defines has to guard itself -- `r if (runs) x else NA` -- or the vignette
# fails to build on a machine without these packages instead of skipping.
has_gstat  <- requireNamespace("gstat",   quietly = TRUE)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
runs <- has_gstat && has_ggplot

knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  out.width = "100%",
  eval      = runs
)
```

```{r missing-deps, echo=FALSE, eval=!runs, results='asis'}
cat("**gstat or ggplot2 is not installed, so the code below is shown but not",
    "run.** Install them with `install.packages(c(\"gstat\", \"ggplot2\"))`",
    "and rebuild.\n")
```

*This article needs two optional packages: **gstat**, which fits the variogram
behind the autocorrelation range, and **ggplot2** for the figures. When either
is missing the code is shown but not run, and a note at the top says so.*

## The problem with a random fold

A random fold puts a test point's neighbours in the training set. When the field
is spatially autocorrelated, those neighbours carry the test point's value, and
a model with any capacity to memorise location will look up the answer. The
cross-validation score then measures interpolation between nearby training
points, and reports it as a prediction error for ground the model has never
seen.

Everything else in this vignette follows from wanting a number that does not do
that.

## A learner you can read

The effect is easiest to see with a model whose memorising is obvious.
`new_spatial_fit()` is the extension point for putting your own learner on the
package's folds, and five-nearest-neighbour smoothing is about the simplest
thing that fits through it.

```{r learner}
library(sf)
library(spatialkit)

knn_fit <- function(train_sf, response_var = "z", k = 5L) {
  new_spatial_fit(
    subclass       = "knnsurf_fit",
    engine         = list(xy = st_coordinates(train_sf),
                          y = st_drop_geometry(train_sf)[[response_var]],
                          k = k),
    formula        = as.formula(paste(response_var, "~ 1")),
    response_var   = response_var,
    predictor_vars = character(0),
    data_sf        = train_sf
  )
}

predict.knnsurf_fit <- function(object, newdata = NULL, ...) {
  e  <- object$engine
  xy <- if (is.null(newdata)) e$xy else st_coordinates(newdata)
  vapply(seq_len(nrow(xy)), function(i) {
    d <- sqrt((e$xy[, 1] - xy[i, 1])^2 + (e$xy[, 2] - xy[i, 2])^2)
    mean(e$y[order(d)[seq_len(min(e$k, length(d)))]])
  }, numeric(1))
}
residuals.knnsurf_fit <- function(object, ...) object$engine$y - predict(object)
fitted.knnsurf_fit    <- function(object, ...) predict(object)

registerS3method("predict",   "knnsurf_fit", predict.knnsurf_fit)
registerS3method("residuals", "knnsurf_fit", residuals.knnsurf_fit)
registerS3method("fitted",    "knnsurf_fit", fitted.knnsurf_fit)

knn_fn <- function(train_sf, ...) knn_fit(train_sf, "z", k = 5L)
```

Three methods and a constructor. `cv_spatial()` asks nothing else of a learner,
and `?new_spatial_fit` documents what each backend in the package supplies
beyond this minimum.

A simulated exponential field to run it on, with an effective range of 180 units
over a 1000-unit square:

```{r fixture}
set.seed(7)
n  <- 400
xy <- data.frame(x = 5e5 + runif(n, 0, 1000), y = 5e6 + runif(n, 0, 1000))
D  <- as.matrix(dist(xy))
xy$z <- as.numeric(t(chol(exp(-D / 60) + diag(1e-8, n))) %*% rnorm(n)) +
        rnorm(n, sd = 0.3)
pts  <- st_as_sf(xy, coords = c("x", "y"), crs = 32632)
```

## The size of the effect

Same data, same learner, two fold schemes.

```{r the-gap}
# The field was simulated with an effective range of 180 units, so 200-unit
# blocks hold out whole patches of it.  Sizing blocks from the data, when the
# range is not known, is the next section.
fr <- make_folds(pts, k = 5, method = "random_kfold", seed = 1)
fb <- make_folds(pts, k = 5, method = "block_kfold", block_size = 200, seed = 1)

cv_random  <- cv_spatial(pts, "z", character(0), fit_fn = knn_fn, folds = fr)
cv_blocked <- cv_spatial(pts, "z", character(0), fit_fn = knn_fn, folds = fb)

c(random  = cv_random$overall$RMSE,
  blocked = cv_blocked$overall$RMSE,
  ratio   = cv_blocked$overall$RMSE / cv_random$overall$RMSE)
```

```{r gap-text, echo=FALSE, results='asis'}
r <- cv_blocked$overall$RMSE / cv_random$overall$RMSE
cat(sprintf(paste("Blocked folds put the error %.0f%% higher on this fixture.",
                  "Both numbers come from the same 400 observations and the",
                  "same smoother; the difference is entirely in which rows",
                  "were allowed to train on which.\n"), 100 * (r - 1)))
```

Neither number is the truth. The blocked one answers a harder question, and it
is the question most maps are used for: what happens at a location with no
observation nearby.

## The five fold schemes

```{r methods-table, echo=FALSE, results='asis'}
cat(
"| `method` | holds out | use it when |\n",
"|---|---|---|\n",
"| `random_kfold` | a random 1/k of rows | you want the optimistic bound, or the data are not autocorrelated |\n",
"| `block_kfold` | contiguous spatial blocks | the general case; size the blocks from the correlation range |\n",
"| `buffered_loo` | one point, and everything within `buffer` of it | n is small and you can afford n fits |\n",
"| `leave_location_out` | every observation from one site, named by `group_var` | you have repeat visits and the site is the unit |\n",
"| `nndm` | one point, and a distance-matched exclusion | you know where you will predict, and want the fold geometry to match it |\n",
sep = "")
```

`plot_folds()` draws the assignment, with the blocks behind it when the scheme
has any:

```{r plot-folds, fig.height=4.5, fig.alt="A square study area divided into a grid of blocks, with the points coloured by fold. Each fold occupies whole blocks rather than being scattered, so held-out points sit away from the points used to train."}
plot_folds(fb, pts)
```

Look at whether any fold is a thin sliver or a single corner. Fold geometry that
looks wrong on the map produces a score that is wrong in a way no summary
statistic will tell you about.

## How wide should a block be?

Wide enough that a test block's points are uncorrelated with the training points
outside it, which makes the autocorrelation range the natural scale.

```{r sac}
sac <- estimate_sac_range(pts, "z")
c(range = as.numeric(sac), nugget = sac_nugget(sac))
```

`sac_nugget()` reads the nugget off the same fitted model. A nugget that is most
of the sill says the field has little spatial structure left to leak, and the
blocked and random scores will be close.

`make_folds(auto_range = TRUE)` runs this estimate and uses it as a minimum
block size:

```{r auto-range}
fa <- make_folds(pts, k = 5, method = "block_kfold", response_var = "z",
                 auto_range = TRUE, seed = 1)
c(block_size = as.numeric(fa$params$block_size),
  sac_range  = as.numeric(fa$params$sac_range))
```

`params$block_size` holds the range object itself, so printing it whole also
prints the directional summary; `as.numeric()` above keeps the output to the two
numbers.

When the variogram identifies no range, `estimate_sac_range()` returns `NA` with
the reason attached, `auto_range` logs that it is falling back to geometric
blocks, and you are back to choosing a size yourself. An `NA` there is a finding
about the data. `?estimate_sac_range` lists the five reasons and what each one
means; `plot()` on the returned object draws the variogram that produced it.

## Sizing it by measurement

`cv_block_size_sweep()` runs the cross-validation at several block sizes and
reports the curve, with a random-fold run as the reference.

```{r sweep}
sw <- cv_block_size_sweep(pts, "z", character(0), fit_fn = knn_fn,
                          k = 4, n_sizes = 5, max_fits = 40,
                          seed = 1, quiet = TRUE)
sw
```

```{r plot-sweep, fig.height=4.5, fig.alt="Cross-validated RMSE against block edge length on a log scale. The error rises steadily as the blocks grow, from about 0.80 at the smallest to about 1.13 at the largest, staying above the dashed random-fold reference throughout. A dotted line marks the estimated autocorrelation range."}
plot(sw)
```

Error climbs away from the random-fold reference as blocks widen, and the
estimated range is marked. The height of the climb is what random folds were
hiding.

Read the curve up to the range, and watch `blocks_used` past it. Once
`blocks_used` approaches `k`, each fold is holding out a large contiguous piece
of the study area, and the model is being asked to extrapolate across it. Error
keeps rising for a second reason at that point, and the number stops being about
leakage.

## Reading the result

`overall` is the pooled score and `fold_metrics` is one row per fold.
`plot_cv_metrics()` shows the spread, which matters because a pooled RMSE built
from one catastrophic fold and four good ones describes neither.

```{r plot-metrics, fig.height=4, fig.alt="RMSE for each of five folds, with point size showing how many rows each held out. The values run from about 0.94 to 1.12 around a dashed pooled line near 1.01, so the pooled number hides a spread of nearly twenty percent."}
plot_cv_metrics(cv_blocked, metric = "RMSE")
```

### Every row and every fold is accounted for

Four fields say what happened to the data on the way through:
`n_folds_attempted` against `n_folds_succeeded`, `n_dropped` (rows the data
lost), `orphan_rows` (rows no fold names) and `n_unknown_ids` (fold rows the
data does not have). On the clean run above nothing went missing:

```{r accounting-clean}
c(attempted = cv_blocked$n_folds_attempted, succeeded = cv_blocked$n_folds_succeeded,
  dropped = cv_blocked$n_dropped, orphans = length(cv_blocked$orphan_rows))
```

So here are two runs where something does.

A learner that fails on one fold:

```{r failing-fold}
flaky <- function(train_sf, ...) {
  if (nrow(train_sf) < 320) stop("not enough training rows for this toy learner")
  knn_fit(train_sf, "z", k = 5L)
}
cv_flaky <- cv_spatial(pts, "z", character(0), fit_fn = flaky, folds = fb)

cv_flaky$fold_status
c(attempted = cv_flaky$n_folds_attempted, succeeded = cv_flaky$n_folds_succeeded)
```

`fold_status` names every fold supplied and what became of it, with the error
text for the ones that failed. `status` is one of `ok`, `error`, `skipped`,
`dropped` or `worker_error`. Before this existed, a partial failure left its
causes in console scrollback, which a script or a knitted document does not keep.

Rows the data lost, and rows no fold names:

```{r accounting}
holed <- pts
holed$z[c(3, 17, 42)] <- NA                       # three rows will be dropped
part  <- make_folds(pts[1:300, ], k = 4, method = "block_kfold", seed = 1)

cv_acc <- cv_spatial(holed, "z", character(0), fit_fn = knn_fn, folds = part)

c(n_dropped     = cv_acc$n_dropped,
  n_orphans     = length(cv_acc$orphan_rows),
  n_unknown_ids = cv_acc$n_unknown_ids)
```

`n_dropped` is how many rows `prep_model_data()` removed for a missing or
non-finite value before any fold was fitted. `orphan_rows` holds the row IDs no
fold names, which here is rows 301 to 400, because the folds were built on the
first 300. `n_unknown_ids` counts the distinct row IDs the folds name that the
data does not have, which here is the three rows the `NA`s removed.

The two mirror each other, and which one is non-zero tells you which way the
mismatch runs:

```{r mirror}
cv_rev <- cv_spatial(pts[1:300, ], "z", character(0), fit_fn = knn_fn, folds = fb)
c(n_orphans = length(cv_rev$orphan_rows), n_unknown_ids = cv_rev$n_unknown_ids)
```

Folds built on more rows than the data has give `n_unknown_ids`; folds built on
fewer give `orphan_rows`. Both are zero when the folds and the data agree, which
is the only case in which the score covers every observation.

## Calibration, for a model with intervals

A Bayesian fit predicts an interval, and the question becomes whether the
intervals are the width they claim. `plot_calibration()` draws nominal coverage
against realised coverage per fold, from a `cv_bayes()` result.

```{r calibration, eval=FALSE}
cvb <- cv_bayes(pts, "z", character(0), folds = fb)   # needs brms
plot_calibration(cvb)
```

Points below the diagonal are intervals that are too narrow: a nominal 90
percent interval containing 70 percent of the held-out values is the model
reporting more confidence than it has earned. This chunk is not run here,
because it needs `brms` and a Stan toolchain.

## Next

`vignette("resolution")` covers choosing a cell count before any of this.
`vignette("spatialkit_nc_demo")` runs the whole pipeline on real boundaries.
`?make_folds` documents every scheme and its arguments, `?cv_spatial` the return
value in full, and `?cv_block_size_sweep` the fit budget that `max_fits` caps.
