---
title: "spatialkit: Tessellations, Spatial Cross-Validation and Models"
author: "Justin Chase"
date: "2026-08-26"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{spatialkit: Tessellations, Spatial Cross-Validation and Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# ggplot2 is a Suggests package, and every chunk below either draws something
# or feeds something that does.  R CMD build runs this file, so the whole
# vignette is gated on ggplot2 being installed rather than calling library()
# unguarded and failing the build on a machine without it.
has_ggplot  <- requireNamespace("ggplot2",  quietly = TRUE)
has_geom    <- requireNamespace("geometry", quietly = TRUE)
has_ranger  <- requireNamespace("ranger",   quietly = TRUE)
has_gstat   <- requireNamespace("gstat",    quietly = TRUE)
has_gwmodel <- requireNamespace("GWmodel",  quietly = TRUE) &&
               requireNamespace("sp",       quietly = TRUE)

knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  out.width  = "100%",
  eval       = has_ggplot
)
```

```{r no-ggplot, echo=FALSE, eval=!has_ggplot, results='asis'}
cat("**ggplot2 is not installed, so the code below is shown but not run.**",
    "Install it with `install.packages(\"ggplot2\")` and rebuild.\n")
```

## Overview

This vignette generates **synthetic spatial data** over North Carolina and
walks through the `spatialkit` workflow end to end:

1. Build four tessellation types (Voronoi, hex, square, Delaunay)
2. Assign points to cells and aggregate with `summarize_by_cell()`
3. Draw choropleths of the cell-level mean response
4. Estimate the autocorrelation range and build spatial CV folds
5. Fit a model, and score it under **blocked** versus **random** folds
6. Predict onto a surface and mark where that surface is extrapolation

Everything is self-contained — the boundary comes from the `nc.shp` demo
shapefile bundled with `sf`, so no external files are needed.

Optional backends are checked at the top and every section that needs one is
guarded, so this document renders with whatever you happen to have installed:

```{r backend-report, echo=FALSE}
data.frame(
  package   = c("ggplot2", "geometry", "ranger", "gstat", "GWmodel + sp"),
  available = c(has_ggplot, has_geom, has_ranger, has_gstat, has_gwmodel),
  used_for  = c("all plots", "true Delaunay triangulation",
                "random forest backend", "variogram / autocorrelation range",
                "GWR backend")
)
```

---

## 1. Packages and boundary

```{r load-packages}
library(spatialkit)
library(sf)
library(dplyr)
library(ggplot2)

set.seed(42)
```

We load the North Carolina county boundaries shipped with `sf`, dissolve them
into a single state outline, and project to **NAD83 / North Carolina (ftUS)**
(EPSG:2264) so distances are planar rather than angular. Every distance,
bandwidth and block size below is therefore in **US survey feet**, not metres —
projected does not mean metric, and the unit is whatever the CRS says it is:

```{r boundary}
nc_counties <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
nc_boundary <- nc_counties |>
  st_union() |>
  st_transform(2264) |>
  st_as_sf()
```

---

## 2. Synthetic observations

300 points inside the state boundary with two predictors and a spatially
varying response:

- `elevation` — gradient increasing west to east, plus noise
- `pop_density` — decays with distance from two fake "cities"
- `y` — driven by the two predictors **plus a spatial field neither of them
  explains**. That last term is deliberate: it is the unmodelled spatial
  structure that makes random cross-validation optimistic, and it is what
  section 5 measures.

```{r fake-data}
n_points <- 300

pts_raw    <- st_sample(nc_boundary, size = n_points, type = "random")
pts_coords <- st_coordinates(pts_raw)
x_coords   <- pts_coords[, 1]
y_coords   <- pts_coords[, 2]

elevation <- scale(x_coords)[, 1] * 500 + rnorm(n_points, 3000, 400)

city1 <- c(1530000, 550000)   # Charlotte-ish in EPSG:2264
city2 <- c(2150000, 750000)   # Raleigh-ish
dist_to_city <- pmin(
  sqrt((x_coords - city1[1])^2 + (y_coords - city1[2])^2),
  sqrt((x_coords - city2[1])^2 + (y_coords - city2[2])^2)
)
pop_density <- pmax(exp(-dist_to_city / 400000) * 5000 + rnorm(n_points, 200, 100), 10)

spatial_field <- 20 * sin(x_coords / 250000) * cos(y_coords / 250000)

y_response <- 50 +
  0.01  * elevation +
  0.005 * pop_density +
  spatial_field +
  rnorm(n_points, 0, 5)

points_sf <- st_sf(
  y           = y_response,
  elevation   = elevation,
  pop_density = pop_density,
  geometry    = pts_raw
)
```

```{r quick-peek, fig.height=3.6}
ggplot() +
  geom_sf(data = nc_boundary, fill = "grey95", colour = "black") +
  geom_sf(data = points_sf, aes(colour = y), size = 1.1) +
  scale_colour_viridis_c(name = "Response (y)") +
  theme_void() +
  ggtitle("Raw observation points, North Carolina")
```

---

## 3. Four tessellations

### 3a. Voronoi from ~40 k-means seeds

Voronoi cells are built around *seed* points, not the observations themselves.
Seeding one cell per observation would give 300 cells each containing a single
point — a nearest-neighbour interpolation rather than an aggregation, with no
within-cell variation to compute a standard error from. `get_voronoi_seeds()`
clusters the observations first, so cell size follows sampling density.

```{r tess-voronoi}
seeds <- get_voronoi_seeds(
  boundary      = nc_boundary,
  sample_points = points_sf,
  method        = "kmeans",
  n             = 40
)

tess_voronoi <- build_tessellation(
  seeds, boundary = nc_boundary,
  method = "voronoi", clip = TRUE, quiet = TRUE
)
```

### 3b and 3c. Hex and square grids (~50 cells)

```{r tess-grids}
tess_hex <- build_tessellation(
  points_sf, boundary = nc_boundary,
  method = "hex", approx_n_cells = 50, clip = TRUE, quiet = TRUE
)

tess_square <- build_tessellation(
  points_sf, boundary = nc_boundary,
  method = "square", approx_n_cells = 50, clip = TRUE, quiet = TRUE
)
```

All four methods return cells carrying a `cell_id` column; the grid methods
keep `poly_id` alongside it, holding the same values.

```{r cell-id-check}
names(tess_hex$cells)
```

### 3d. Delaunay triangles

`method = "triangles"` uses `geometry::delaunayn()` when `geometry` is
installed. Without it, it does **not** error — it falls back to
`sf::st_triangulate()` (GEOS) on the point set and logs a warning. That is
still the Delaunay triangulation of the input points; only the resolution of
degenerate configurations can differ. Because the two paths are not identical,
this section is guarded on `geometry` rather than on the call succeeding:

```{r tess-tri, eval=has_ggplot && has_geom}
tess_tri <- build_tessellation(
  points_sf, boundary = nc_boundary,
  method = "triangles", clip = TRUE, quiet = TRUE
)
nrow(tess_tri$cells)
```

```{r cell-counts}
cat(sprintf(
  "Voronoi: %d | Hex: %d | Square: %d | Triangles: %s\n",
  nrow(tess_voronoi$cells),
  nrow(tess_hex$cells),
  nrow(tess_square$cells),
  if (has_geom) nrow(tess_tri$cells) else "skipped (install 'geometry')"
))
```

---

## 4. Cell-level aggregation with `summarize_by_cell()`

`assign_features_to_polygons()` joins points to cells; `summarize_by_cell()`
does the aggregation — means, standard deviations and standard errors for the
response and every predictor, plus `n` and `cell_weight`. There is no need to
write the `group_by()`/`summarise()` by hand, and doing so loses the
design-effect machinery below.

```{r summarize}
assigned <- assign_features_to_polygons(points_sf, tess_voronoi$cells,
                                        polygon_id_col = "cell_id")

cell_stats <- summarize_by_cell(
  assigned,
  response_var   = "y",
  predictor_vars = c("elevation", "pop_density"),
  id_col         = "cell_id"
)

head(as.data.frame(cell_stats)[, c("cell_id", "n", "resp_mean_y",
                                   "..sd_resp_y", "..se_resp_y")])
```

The `..se_*` columns are **IID** standard errors at the default `deff = 1`,
which is anticonservative when points inside a cell are spatially correlated.
`deff = "kish"` applies Kish's design-effect correction from an estimated
intra-class correlation, and `attr(., "deff_applied")` records what was used:

```{r summarize-deff}
cell_kish <- summarize_by_cell(
  assigned,
  response_var   = "y",
  predictor_vars = c("elevation", "pop_density"),
  id_col         = "cell_id",
  deff           = "kish"
)

deff <- attr(cell_kish, "deff_applied")
cat(sprintf("method = %s | ICC(response) = %.3f | median design effect = %.2f\n",
            deff$method, deff$icc_resp, stats::median(deff$deff, na.rm = TRUE)))

# Inflation of the response standard error, cell by cell.
summary(cell_kish$`..se_resp_y` / cell_stats$`..se_resp_y`)
```

### Choropleths

```{r choropleth-helper}
#' Assign points to cells, summarise, and draw a choropleth.
make_choropleth <- function(tess, boundary, points, title = NULL) {
  cells    <- tess$cells
  assigned <- assign_features_to_polygons(points, cells, polygon_id_col = "cell_id")

  stats_df <- summarize_by_cell(assigned, response_var = "y",
                                id_col = "cell_id")
  cells    <- left_join(cells, as.data.frame(stats_df)[, c("cell_id", "resp_mean_y")],
                        by = "cell_id")

  plot_tessellation_map(
    tessellation_sf = cells,
    boundary        = boundary,
    fill_col        = "resp_mean_y",
    palette         = "viridis",
    tile_alpha      = 0.9,
    outline_col     = "white",
    outline_size    = 0.3,
    boundary_col    = "grey20",
    boundary_size   = 0.8,
    legend_title    = "Mean y",
    title           = title,
    subtitle        = sprintf("%d cells | %d observations",
                              nrow(cells), nrow(points))
  )
}
```

```{r choro-voronoi, fig.cap="Voronoi choropleth"}
make_choropleth(tess_voronoi, nc_boundary, points_sf, "Voronoi tessellation")
```

```{r choro-hex, fig.cap="Hex grid choropleth"}
make_choropleth(tess_hex, nc_boundary, points_sf, "Hexagonal grid")
```

```{r choro-square, fig.cap="Square grid choropleth"}
make_choropleth(tess_square, nc_boundary, points_sf, "Square grid")
```

```{r choro-tri, fig.cap="Delaunay choropleth", eval=has_ggplot && has_geom}
make_choropleth(tess_tri, nc_boundary, points_sf, "Delaunay triangulation")
```

```{r comparison-panel, fig.width=7, fig.height=2.8, fig.cap="All three at a glance", eval=has_ggplot && requireNamespace("patchwork", quietly = TRUE)}
library(patchwork)

(make_choropleth(tess_voronoi, nc_boundary, points_sf, "Voronoi") |
 make_choropleth(tess_hex,     nc_boundary, points_sf, "Hex") |
 make_choropleth(tess_square,  nc_boundary, points_sf, "Square")) +
  plot_annotation(title = "Tessellation comparison, cell-level mean response")
```

---

## 5. Spatial cross-validation

This is the part the rest of the package exists to support.

### 5a. How far does correlation reach?

`estimate_sac_range()` fits directional variograms and returns the effective
range in CRS units — here, US survey feet. It returns `NA` (still classed
`sac_range`, so it prints as a bare `NA`) when the empirical variogram never
reaches a sill, because an unidentified range must not be used to size blocks.

```{r sac-range, eval=has_ggplot && has_gstat}
sac <- estimate_sac_range(points_sf, response_var = "y",
                          predictor_vars = c("elevation", "pop_density"))
sac

if (is.na(sac)) {
  cat("no identified range:", attr(sac, "rejected_reason"), "\n")
} else {
  cat(sprintf("range = %.0f ft; anisotropy ratio %.2f\n",
              as.numeric(sac), attr(sac, "anisotropy")))
}
```

The fit is attached either way, so `plot(fit, type = "variogram")` on a fitted
model can draw the curve and let you judge it rather than trust it. A rejected
range must not be handed to `make_folds(auto_range = TRUE)`, which is why it
comes back `NA` rather than as a long range.

### 5b. Two fold schemes on the same data

```{r folds}
folds_random  <- make_folds(points_sf, k = 5, method = "random_kfold", seed = 42)
folds_blocked <- make_folds(points_sf, k = 5, method = "block_kfold",  seed = 42)

c(random = folds_random$k, blocked = folds_blocked$k)
```

`plot_folds()` is the fastest way to see whether the blocks actually separate
the data or are smaller than the autocorrelation range and therefore leaking:

```{r plot-folds, fig.width=7, fig.height=3, eval=has_ggplot && requireNamespace("patchwork", quietly = TRUE)}
library(patchwork)
plot_folds(folds_random,  points_sf, boundary = nc_boundary) +
  plot_folds(folds_blocked, points_sf, boundary = nc_boundary)
```

### 5c. The number the fold scheme changes

`fit_rf_model()` and `cv_rf()` need `ranger`. `include_coords = TRUE` hands the
forest the coordinates, which lets it memorise the training surface — the
failure mode the default (`FALSE`) exists to prevent, and the one random folds
cannot see. Scoring the same model on both fold schemes shows the size of the
gap:

```{r cv-contrast, eval=has_ggplot && has_ranger}
rf_args <- list(include_coords = TRUE, num_trees = 300, seed = 1)

cv_random  <- do.call(cv_rf, c(list(points_sf, "y", c("elevation", "pop_density"),
                                    folds = folds_random),  rf_args))
cv_blocked <- do.call(cv_rf, c(list(points_sf, "y", c("elevation", "pop_density"),
                                    folds = folds_blocked), rf_args))

data.frame(
  folds = c("random_kfold", "block_kfold"),
  R2    = c(cv_random$overall$R2,   cv_blocked$overall$R2),
  RMSE  = c(cv_random$overall$RMSE, cv_blocked$overall$RMSE)
)
```

```{r cv-contrast-text, echo=FALSE, results='asis', eval=has_ggplot && has_ranger}
cat(sprintf(
  "Random folds report R<sup>2</sup> = %.3f; blocked folds report %.3f on the same fitted model — a drop of %.0f%% of the reported skill.\n",
  cv_random$overall$R2, cv_blocked$overall$R2,
  100 * (cv_random$overall$R2 - cv_blocked$overall$R2) / cv_random$overall$R2))
```

The blocked estimate is the one to report. The random one describes
interpolation between points you already have, which is not the task.

---

## 6. Fitting a model

```{r rf-fit, eval=has_ggplot && has_ranger}
rf_fit <- fit_rf_model(points_sf, response_var = "y",
                       predictor_vars = c("elevation", "pop_density"))
rf_fit
```

Note that `summary()` on an `rf_fit` reports **out-of-bag** metrics, not
in-sample ones — `fitted.rf_fit()` returns out-of-bag predictions, and the
printout says so. A `gwr_fit` or `bayesian_fit` reports genuinely in-sample
metrics, so the two are not comparable; `compare_models_cv()` is.

```{r rf-summary, eval=has_ggplot && has_ranger}
summary(rf_fit)
rf_fit$info$importance     # coef() on an rf_fit errors: a forest has no coefficients
```

### Residual diagnostics

`plot()` on any `spatial_fit` maps the residuals; visible structure means
unmodelled spatial autocorrelation. `residual_morans_i()` puts a number on it.

```{r rf-resid, fig.height=3.6, eval=has_ggplot && has_ranger}
plot(rf_fit, type = "residuals")
```

```{r rf-moran, eval=has_ggplot && has_ranger}
mi <- residual_morans_i(rf_fit)
cat(sprintf("Moran's I = %.4f (z = %.2f, p = %.3g)\n",
            mi$observed, mi$z, mi$p_value))
```

### GWR, if `GWmodel` is installed

```{r gwr-fit, eval=has_ggplot && has_gwmodel}
gwr_fit <- fit_gwr_model(
  data_sf        = points_sf,
  response_var   = "y",
  predictor_vars = c("elevation", "pop_density"),
  adaptive       = TRUE,
  kernel         = "bisquare"
)

gwr_met <- model_metrics(gwr_fit)
cat(sprintf("Bandwidth: %.1f | in-sample R2: %.3f | RMSE: %.3f\n",
            gwr_fit$info$bandwidth, gwr_met$R2, gwr_met$RMSE))
```

### Comparing backends on identical folds

`compare_models_cv()` cross-validates each requested backend on the `folds` you
pass. Any backend whose package is missing is dropped with a message, so the
call still returns whatever could run.

```{r compare, eval=has_ggplot && has_ranger}
cmp <- compare_models_cv(
  points_sf, "y", c("elevation", "pop_density"),
  models   = c("RF", "GWR"),
  folds    = folds_blocked,
  rf_args  = list(num_trees = 300)
)
cmp$overall
```

---

## 7. From a fit to a map, and where the map applies

`predict_surface()` builds a regular grid over the training extent, joins
covariates from the nearest observation, clips to the boundary and predicts:

```{r surface, fig.height=3.6, eval=has_ggplot && has_ranger}
surf <- predict_surface(rf_fit, n_cells = 3000,
                        covariates = points_sf, boundary = nc_boundary)

ggplot() +
  geom_sf(data = surf, aes(colour = .pred), size = 0.6) +
  geom_sf(data = nc_boundary, fill = NA, colour = "grey20") +
  scale_colour_viridis_c(name = "Predicted y") +
  theme_void() +
  ggtitle("Predicted surface")
```

A fitted model returns a number for any location you hand it, including
locations whose predictor values look nothing like the training data.
`area_of_applicability()` marks where the cross-validated score actually
applies. Pass the folds you validated with:

```{r aoa, fig.height=3.6, eval=has_ggplot && has_ranger}
aoa <- area_of_applicability(surf, model = rf_fit, folds = folds_blocked)

cat(sprintf("inside: %d | outside: %d | undetermined: %d | DI threshold %.3f\n",
            aoa$n_inside, aoa$n_outside, aoa$n_na, aoa$threshold))

# AOA is NA wherever a predictor was missing or non-finite, and `!NA` is NA,
# which R silently skips in a subscripted assignment. Test for TRUE and let
# anything else count as outside.
inside <- aoa$aoa$AOA %in% TRUE
surf$.pred_masked <- ifelse(inside, surf$.pred, NA_real_)

ggplot() +
  geom_sf(data = surf, aes(colour = .pred_masked), size = 0.6) +
  geom_sf(data = nc_boundary, fill = NA, colour = "grey20") +
  scale_colour_viridis_c(name = "Predicted y", na.value = "grey85") +
  theme_void() +
  ggtitle("Predicted surface, extrapolations blanked out")
```

Nothing is masked here, and the reason is worth understanding rather than
taking as reassurance: `predict_surface(covariates = points_sf)` copies
predictor values from the **nearest observation**, so every grid cell carries a
predictor vector some training point already had. Its dissimilarity index is
therefore near zero by construction. The index earns its keep when the
covariates come from somewhere else — a raster, a different survey, a future
scenario. Hand it values outside the training range and it says so:

```{r aoa-outside, eval=has_ggplot && has_ranger}
extreme <- st_sf(
  elevation   = c(mean(points_sf$elevation), max(points_sf$elevation) * 4),
  pop_density = c(mean(points_sf$pop_density), max(points_sf$pop_density) * 4),
  geometry    = st_geometry(points_sf)[1:2]
)
area_of_applicability(extreme, model = rf_fit, folds = folds_blocked)$aoa[, c("DI", "AOA")]
```

---

## Summary

| Tessellation | Cells | Notes |
|:---|---:|:---|
| Voronoi | `r if (has_ggplot) nrow(tess_voronoi$cells) else NA` | Adapts to point density via k-means seeds |
| Hex grid | `r if (has_ggplot) nrow(tess_hex$cells) else NA` | Uniform hexagons, good for regular sampling |
| Square grid | `r if (has_ggplot) nrow(tess_square$cells) else NA` | Simplest regular grid |
| Delaunay | `r if (has_ggplot && has_geom) nrow(tess_tri$cells) else NA` | One triangle per Delaunay triplet; finest resolution |

The choropleths show how each tessellation aggregates the response. Section 5
shows the thing that matters most: on the same data and the same fitted model,
the fold scheme moves the reported score substantially, and only the blocked
number describes the task you actually have.

```{r session-info, echo=FALSE, eval=TRUE}
sessionInfo()
```
